Extended Data

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
references

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
However, model representation of C allocation to stem growth is overly simplistic in that it does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@friend_need_2019].
As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@friend_need_2019]. 
Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022]. 
In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@ref], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

>>>>

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020;@fatichi_moving_2014].

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022]. 
In contrast, little is known about broadleaf deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.
